---
title: "Exercise 1 --- Basic Maps and Graphs in R"
subtitle: "BSTA 374 · 12 points · Due Thursday, September 10, 2026, 11:59 pm"
author: "Elizabeth Ehrhardt · Lehigh University"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
  pdf:
    papersize: letter
    geometry:
      - margin=1in
    toc: false
    colorlinks: true
    code-overflow: wrap
prefer-html: true
execute:
  eval: false
  echo: true
---

## What this exercise is

Your first graded work in this course. It covers the first two weeks: reading
spatial data with `sf`, understanding what a coordinate reference system is and
moving between them, and producing honest choropleth and point maps with both
`ggplot2` and `tmap`.

**No data to download.** Everything runs on the North Carolina SIDS data that
ships inside the `sf` package, so your analysis will reproduce on my machine
exactly as it does on yours --- which is the point.

Each part ends with **starting code**. Copy it, fill in the blanks marked
`______`, and adapt it. It is scaffolding, not a solution: the mechanical lines
are given so that your time goes into the parts that require judgement.

::: {.callout-important}
## Submission
Submit **one file** to Course Site:

- `LASTNAME_FI_Exercise1.qmd` --- the source

Example: `DELMELLE_EM_Exercise1.qmd`. Put your name in the document header.

**Due Thursday, September 10, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Before you submit: restart R (`Ctrl/Cmd + Shift + F10`) and render the document
from a clean session. If it does not run start to finish, it is incomplete ---
regardless of how good the maps are. This is the habit that matters most, and
we start enforcing it now.
:::

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(tmap)
# add anything else you use -- and load it here, not halfway down

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

---

## Part A --- Read and describe the data (2 points)

Answer each in **one or two sentences of prose**, outside the code chunks, and
show the code you used to find out.

1. How many features does `nc` contain, and what does one row represent?
nc contains 15 different features. One row represents one county in North Carolina. 

2. What geometry type is stored? Why is it that type rather than `POLYGON`?
   *(Look at the coast.)*
The geometry type stored is multipolygon. It is a multipolgyon rather than a polygon because some of the North Carolina counties are islands, and are a separate entity from the inland counties. 
   
3. What coordinate reference system is the data in? Give its name, say whether it
   is geographic or projected, and say how you found out.
The data uses NAD83 (North American Datum 1983), which is an unprojected Geographic Coordinate System. I found out by looking at the range of the numberse on the dataframe, like how far the decimals went, and then used Gemini to help me narrow down the different systems. 
   
4. In your own words, what do `BIR74` and `SID74` measure? Why does having both
   matter more than having either one alone?
BIR74 measures the county's births, SID74 measures the county's sudden infant death rate. Having both matters because if a county with 100 people has 5 SID cases, it is less of an issue than a county that has 1,000 birthds. The prevalence of SIDs is based off of the birth rate. 

### Starting code

```{r}
#| label: a-starter

# A1 -- how big is it, and what are the columns?
nrow(nc)
names(nc)
nc                       # the header block above the table says a lot

# A2 -- one geometry type per feature; count them
table(st_geometry_type(nc))

# A3 -- three ways of asking about the CRS
st_crs(nc)               # the full definition
st_crs(nc)$input         # the short name
st_is_longlat(nc)        # TRUE = geographic (degrees), FALSE = projected (metres)

# A4 -- the help page documents every column
?nc
```

---

## Part B --- Coordinate reference systems (2 points)

1. Transform `nc` to **EPSG:32119** (NAD83 / North Carolina State Plane, in
   metres) and store it as a **new object**. Do not overwrite `nc` --- you need
   both.
2. Plot the counties **before and after**, near enough to each other that a reader
   can compare them. One or two sentences on what visibly changed.
   The state appears with more curvature as if it was on a 3D projection rather than on a 2D. 
   
3. Compute the area of every county in **square kilometres**, using the projected
   version. Report the five largest.
   The five largest counties are Columbus, Robeson, Sampson, Bladen, and Wake. 
   
4. Now compute areas from the *unprojected* object as well. Compare the two sets
   of numbers and explain, in two or three sentences, what `sf` did and why the
   projected version is the one you would report in a paper.
  
Sf computes the area on a curved surface, whereas not using sf computes the area on a 2D surface. Sf includes calculations that account for curvature and longitudinal convergence, whereas non sf uses planar values for calculations. The projected version is the one you would report in a paper because it is more accurate to local scale and area distortion in North Carolina- making for more standard calculations. 

::: {.callout-note}
Step 4 will not give you an error, and the two sets of numbers will be close but
not identical. That is the interesting part --- the question is which *surface*
each answer refers to.
:::

### Starting code

```{r}
#| label: b-starter

# B1
nc_proj <- st_transform(nc, 32119)

# B2 -- one plot each
p1 <- ggplot(nc)      + geom_sf(fill = "grey90", colour = "white") + theme_void()
p2 <- ggplot(nc_proj) + geom_sf(fill = "grey90", colour = "white") + theme_void()

p1
p2
# side by side, if you like:  library(patchwork);  p1 + p2

# B3 -- st_area() returns square METRES, as a units object
nc_proj <- nc_proj |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc_proj), km^2)))

nc_proj |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)

# B4 -- now do the same starting from nc (degrees) and compare
nc <- nc |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc), km^2)))

nc |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)
```

---

## Part C --- Choropleth maps (5 points)

This is the largest part of the exercise, and it is as much about judgement as
about code.

1. Add a column giving the **SIDS rate per 1,000 live births** for 1974--78.

2. Make a choropleth of the **raw counts** (`SID74`) and one of the **rate**,
   using `ggplot2`. Then write three or four sentences explaining why the two maps
   disagree, and which one you would put in front of a health department, and why.
   The two maps disagree because one is displaying the raw counts of SIDs and one is looking at the rate adjusted for how many people are in each county. In terms of the darker shaded county, the maps tell different story. The rate map shows Union county having the highest rates of SIDS, despite the other map showing that Mecklenburg has the highest counts. If I was going to present these maps in front of a health department to pitch an intervention, I would show the rates map. This is because it puts more into perspective the scale of the death rate in comparison to the whole population. 
   
   
3. Rebuild the rate map using **`tmap`**. Same information, other package.

4. Vary the **class intervals** on the rate map: produce it at least three ways
   --- for instance equal intervals, quantiles, and Jenks natural breaks. Two or
   three sentences on how much the apparent geography of risk changes. Name the
   classification you would defend, and say why. 
   The equal interval map creates classes that are perfectly equal, however since most of the data clusters at lower rates, most of the counties fall into lower categories that doesn't show the true diversity. The quantile map put equal number of counties into each class, so each color section of the map has the same amount of data attributed to it. The Jenks classification breaks the counties into gaps that actually exist, so actual risk and high rates are able to be distinguished using this model. For this reason, I would defend the Jenks classification for its ability to handle skewed data and detail true rate distributions. 
   
5. Every map you submit must have a **legend with units**, a **title**, thin
   boundaries that do not overwhelm the fill, and a **perceptually uniform,
   colour-blind-safe palette**. State in one sentence which palette you chose and
   why, including whether dark means high or low. I chose the mako pallette because it was visually appealling, I could tell where the breaks were, and because there was a large darkness difference in between the "high" values and the "low" values. In mako, dark means high and light means low, which felt like common sense for what these maps are conveying. 

::: {.callout-tip}
## What earns full marks here
Not the prettiest map --- the map whose choices you can defend. A student who
picks quantiles and explains the trade-off in two clear sentences scores higher
than one who picks Jenks because it looked nicer.
:::

### Starting code

```{r}
#| label: c-starter

# C1 -- deaths per 1,000 live births
nc <- nc |>
  mutate(rate74 = 1000 * SID74 / BIR74)

# C2 -- counts, then rates. Same code, one variable changed.
countmap <- ggplot(nc) +
  geom_sf(aes(fill = SID74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "Count") +
  labs(title = "SIDS count, 1974-1979") +
  theme_void()

ratemap <- ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(option = "mako", direction = -1, name = "Per 1,000") +
  labs(title = "SIDS rate, 1974-1979") +
  theme_void()

countmap
ratemap

# other palettes to try:
#   scale_fill_viridis_c(option = "viridis" / "plasma" / "cividis" / "mako")
#   scale_fill_distiller(palette = "YlOrRd", direction = 1)
#   RColorBrewer::display.brewer.all()   # shows every ColorBrewer option

# C3 -- the same map in tmap
tmap_mode("plot")                        # "view" makes it interactive

tm_shape(nc) +
  tm_polygons(fill = "rate74") +
  tm_title("SIDS rate, 1974-1979")

# C4 -- change ONLY the classification, so the comparison is fair
tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_title("Quantiles")

tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "equal", n = 5)) +
  tm_title("Equal")

tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "jenks", n = 5)) +
  tm_title("Jenks")

# then repeat with style = "equal" and style = "jenks"
# other options: "pretty", "fisher"
```

::: {.callout-warning}
## A note on `tmap` versions
`tmap` changed its argument names in version 4. If an example you find online
uses `tm_polygons(col = ...)` or `style = ...` directly and yours errors, check
`packageVersion("tmap")` and read the help page rather than guessing. The code
above is written for version 4.
:::

---

## Part D --- A point layer (2 points)

1. Look up the approximate longitude and latitude of **four cities in North
   Carolina** and build a small table holding the name and the two coordinates.
   Cite where you got the coordinates. 
   I got the coordinates from Claude AI.
2. Turn that table into an `sf` object with the correct CRS. State which EPSG
   code you used and why. 
   I chose EPSG 4326 because it is the reference system that expresses locations as unprojected coordinate pairs in decimal degrees. 
3. Plot the four cities **on top of** your rate choropleth, labelled legibly.
4. One sentence: what would go wrong if you built the `sf` object with the
   coordinate columns in the other order?

### Starting code

```{r}
#| label: d-starter

# D1 -- North Carolina longitudes are NEGATIVE (west of Greenwich)
cities <- tibble(
  name = c("Swansboro", "New Bern", "Asheville", "Greensboro"),
  lon  = c(-77.12, -77.04, -82.55, -79.79),
  lat  = c(34.69, 35.11, 35.60, 36.07)
)

# D2 -- coords takes the column names in x, y order: longitude FIRST
cities_sf <- cities |>
  st_as_sf(coords = c("lon", "lat"), crs = 4326)

# D3 -- layers draw in order, so points come after polygons
ratemap +
  geom_sf(data = cities_sf, size = 2.5, colour = "black") +
  geom_sf_text(data = cities_sf, aes(label = name), size = 3, nudge_y = 0.15)

# check your cities landed in North Carolina and not in the ocean
```

---

## Part E --- Reproducibility and AI use (1 point)

1. Your document renders from a clean session, with every package loaded
   explicitly. *(Check this last, after you have stopped editing.)*
2. End the document with an **AI use note**: which tools you used, what for, and
   at which stage --- planning, coding, debugging, interpreting, or writing. Two
   or three sentences.
   
   I used claude AI to help me debug question D3. I had incorrectly created my cities_sf tibble with swapped longitude and latitude, and it helped me figure out the source of my problem to create my map. Gemini was also used on A3 to help me figure out the coordinate reference system the data was in. 

Remember the course policy: AI use is unrestricted and costs you nothing, but you
must be able to explain every line you hand in. Disclosure is the only
requirement.

### Starting code

```{r}
#| label: e-starter

sessionInfo()
```

---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct inspection of the object; clear, accurate answers | 2 |
| **B** | Correct transformation and areas; comparison explained | 2 |
| **C** | Both packages working; counts-vs-rates argument; class intervals compared and a defended choice; cartographic standards met | 5 |
| **D** | Point layer built with the right CRS and correctly overlaid | 2 |
| **E** | Renders from a clean session; AI use note present | 1 |
| | **Total** | **12** |

Writing counts for roughly 10% of the mark within each part: the prose should be
in complete sentences and free of obvious errors. Every part asks for
interpretation as well as code --- a submission that is all code and no sentences
cannot score above about 60%.

---

## Getting help

- Office hours, or email me --- with the code and the exact error message, not a
  screenshot of the whole console.
- The `sf` documentation is genuinely good: `vignette("sf1", package = "sf")`.
- Moraga Chapter 2 covers most of Parts A--C.
- If you are stuck for more than about thirty minutes on something that feels
  mechanical, that is a signal to ask rather than to persevere.